---
title: "CES fragment data analysis"
author: "Eugenio Graceffo"
date: "July 2024"
output:
  html_document:
    toc: True
    toc_float: True
    df_print: paged
params:
  met_peak: 128
  unmet_peak: 386
  peak_tolerance: 7
  scaling_peak: 200
  reference_sample: "M10U0"
---
## 
What is the lower limit of DNA input to be able to analyse methylation sites by fragment analysis?

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com/"))  # Replace with your preferred CRAN mirror
```


## Package loading
```{r pkg-load, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, results='hide'}


#load libraries
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("readr")) install.packages("readr")
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("readxl")) install.packages("readxl")
if (!require("shiny")) install.packages("shiny")



library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(readxl)
library(shiny)


```



## Retrieve variables from bash GUI
```{r variable-load}
dir.create("Figures")
dir.create("Figures/Statistics")



met_peak <- params$met_peak
unmet_peak <- params$unmet_peak
peak_tolerance <- params$peak_tolerance
scaling_peak <- params$scaling_peak
reference_sample <- params$reference_sample
```


```{r}
## create a vector to include the scaling peak range as a string of characters
scaling_peak_vector <- c(as.character(scaling_peak-0.2),as.character(scaling_peak-0.2), as.character(scaling_peak-0.1), as.character(scaling_peak), as.character(scaling_peak+0.1), as.character(scaling_peak+0.2))

scaling_peak_vector
```





## load peaktable output
```{r data-load}
PeakTable <- read_excel("PeakTable.xlsx", 
    skip = 12)

## wrangle the data
df <- PeakTable %>%
  filter(Dye != is.na(PeakTable$Dye)) %>%
  filter(Sample != "Sample") %>%  ## removes lines with repeated column names
  filter(Dye == "Blue") %>%
  select(Sample, Size, Height, Area, Score) %>%
  separate(Sample, into = c("V1", "Code"), sep = "_", extra = "drop", remove = FALSE) %>%
  select(-"V1")



df_scaling_factor <- PeakTable %>%
  filter(Dye != is.na(PeakTable$Dye)) %>%
  filter(Sample != "Sample") %>%  ## removes lines with repeated column names
  filter(Dye == "Orange", Size %in% scaling_peak_vector) %>%
  select(Sample, Size, Height, Area, Score) %>%
  separate(Sample, into = c("V1", "Code"), sep = "_", extra = "drop", remove = FALSE) %>%
  select(-"V1") %>%
  select(-Size)

df <- df %>%
  full_join(df_scaling_factor, suffix = c("", "_scaling_value"), by=c("Sample","Code"))
  

## Adjust column types
df$Size <- as.double(df$Size)
df$Height <- as.double(df$Height)
df$Area <- as.double(df$Area)
df$Score <- as.double(df$Score)

df$Height_scaling_value <- as.double(df$Height_scaling_value)
df$Area_scaling_value <- as.double(df$Area_scaling_value)
df$Score_scaling_value <- as.double(df$Score_scaling_value)


PeakTable
df
```





## Peaks calculations (+- peak tollerance)
Calculate the Area Under the Curve (AUC) as the sum of peaks +- peak tollerance 
met -> 128 bp
unmet -> 386 bp
```{r peak-assign}
## Assign peak type
df$Peak_type <- ifelse(df$Size >(met_peak-peak_tolerance) & df$Size <(met_peak+peak_tolerance) , "met", 
                  ifelse(df$Size >(unmet_peak-peak_tolerance) & df$Size <(unmet_peak+peak_tolerance) , "Unmet", "Other"))

df <- relocate(df, "Peak_type", .after = "Size")

df
```
Calculate normalized values of Height and AUC
```{r normalization}
df <- df %>%
  mutate(Height_normalized = Height/Height_scaling_value) %>%
  mutate(Area_normalized = Area/Area_scaling_value)

df

```

Calculate the sum of the AUC and select the highest of the heights
```{r AUC-sum_Height-selection}
df <- df %>%
  group_by(Code, Peak_type) %>%
  mutate(AUC_sum = sum(Area_normalized)) %>%
  mutate(Height_max = max(Height_normalized), .after = Height_normalized)

```




## load metadata
```{r metadata-load}
metadata <- read_excel("Metadata.xlsx")

## remove blank wells
metadata <- metadata %>%
  filter(Replicate_name != is.na(metadata$Replicate_name))

metadata
```
```{r data-merge}
## merge
df <- df %>%
  full_join(metadata) %>%
  relocate(c(Replicate_name, Sample_name, Condition), .after = "Code")

df
```



### Calculations

```{r summary-values-calculation}

## separate dataframe for filtering and plotting
df_plot <- df %>%
  filter(Peak_type != "Other") %>%
  select(-Size, -Height, -Area, -Score, -Height_scaling_value, -Area_scaling_value, -Score_scaling_value, -Height_normalized, -Area_normalized)%>%
  unique() %>%
  separate(Sample, into = c("V1", "temp"), sep = "-", extra = "drop", remove = FALSE) %>%
  separate(temp, into = c("order"), sep = "_", extra = "drop", remove = TRUE) %>%
  select(-"V1") %>%
  arrange(as.double(order))
df_plot$order <- as.double(df_plot$order)


## create a dataframe for the ordering in the figures
order_df <- df_plot %>%
  ungroup()%>%
  select(order, Sample_name) %>%
  group_by(Sample_name) %>%
  summarise(order = mean(order)) %>%
  unique()

## calculate the percentage per each replicate
 df_plot <- df_plot %>%
  group_by(Code) %>%
  mutate(AUC_sum_percentage_rep = AUC_sum / sum(AUC_sum)) %>%
  mutate(Height_max_percentage_rep = Height_max / sum(Height_max))
 
 df_plot

## calculate the mean per each replicate and THEN the percentage per condition
 df_plot_percentages_mean <- df_plot %>%
  group_by(Sample_name, Condition, Peak_type) %>%
  mutate(AUC_mean = mean(AUC_sum)) %>%
  mutate(Height_max_mean = mean(Height_max)) %>%
  select(Sample_name, Condition, Peak_type, AUC_mean, Height_max_mean) %>%
  unique() %>%
  left_join(order_df)%>%
  arrange(Peak_type)

 df_plot_percentages_mean <- df_plot_percentages_mean %>%
  group_by(Sample_name, Condition) %>%
  mutate(AUC_sum_percentage_average = round(AUC_mean / sum(AUC_mean), digits = 2)) %>%
  mutate(Height_max_percentage_average = round(Height_max_mean / sum(Height_max_mean), digits = 2))
 
 df_plot_percentages_mean
 
 
 ## generate final output table
 
 output_df <- df %>%
   full_join(df_plot) %>%
   full_join(df_plot_percentages_mean)
 
 write_excel_csv(output_df, "Results.csv")
 
 output_df
```

### barplots
```{r barplots-summary}
## mean values plotted
 
 ggbarplot(df_plot_percentages_mean, "Sample_name", "AUC_mean",
  fill = "Peak_type", color = "Peak_type", palette = "Paired") +
    facet_wrap(~Condition)+ 
    ylab("Area Under Curve (AUC) mean")+
    geom_text(aes(label = AUC_sum_percentage_average),
              position = position_stack(vjust = 0.5))

ggsave("barplot_AUC_mean.png", path = "./Figures")

## percentages values plotted

 ggbarplot(df_plot_percentages_mean, "Sample_name", "AUC_sum_percentage_average",
  fill = "Peak_type", color = "Peak_type", palette = "Paired") +
    facet_wrap(~Condition)+ 
    ylab("Area Under Curve (AUC) in %")+
    geom_text(aes(label = AUC_sum_percentage_average),
              position = position_stack(vjust = 0.5))
 
 ggsave("barplot_AUC_percentage.png", path = "./Figures")





ggbarplot(df_plot_percentages_mean, "Sample_name", "Height_max_mean",
  fill = "Peak_type", color = "Peak_type", palette = "Paired") +
    facet_wrap(~Condition)+ 
    ylab("Max Height mean")+
    geom_text(aes(label = Height_max_percentage_average),
              position = position_stack(vjust = 0.5))

ggsave("barplot_Height_max_mean.png", path = "./Figures")


ggbarplot(df_plot_percentages_mean, "Sample_name", "Height_max_percentage_average",
  fill = "Peak_type", color = "Peak_type", palette = "Paired", position = position_stack()) +
    facet_wrap(~Condition)+ 
    ylab("Max Height in %")+
    geom_text(aes(label = Height_max_percentage_average),
              position = position_stack(vjust = 0.5))

ggsave("barplot_Height_max_percentage.png", path = "./Figures")



```







## statistical comparison
```{r barplots-statistics}

for (x in unique(df$Condition)) {
  df_plot_tumor <- df_plot %>%
  filter(Condition == x, Sample_name != 0.01)  ## remove contaminated sample 0.01 and filter for control only



p1 <- ggbarplot(df_plot_tumor, "Sample_name", "AUC_sum",
       add = c("mean_sd"),fill = "Peak_type", color = "black", palette = "Paired", facet.by = "Peak_type") +
    ylab("Area Under Curve (AUC)")+
  guides(x =  guide_axis(angle = 45)) +
  ggtitle(paste0(x, " condition"), subtitle = "t-test") +
  stat_compare_means(label = "p.signif", method = "t.test", ref.group = as.character(reference_sample))

print(p1)
ggsave(paste0(x, "_barplot_AUC_mean_statistics.png"), plot = p1, path = "./Figures/Statistics")

p2 <- ggbarplot(df_plot_tumor, "Sample_name", "AUC_sum_percentage_rep",
       add = c("mean_sd"),fill = "Peak_type", color = "black", palette = "Paired", facet.by = "Peak_type") +
    ylab("Area Under Curve (AUC) in %")+
  guides(x =  guide_axis(angle = 45)) +
  ggtitle(paste0(x, " condition"), subtitle = "t-test") +
  stat_compare_means(label = "p.signif", method = "t.test", ref.group = as.character(reference_sample))

print(p2)
ggsave(paste0(x, "_barplot_AUC_percentage_statistics.png"), plot = p2, path = "./Figures/Statistics")



p3 <- ggbarplot(df_plot_tumor, "Sample_name", "Height_max", 
       add = c("mean_sd"),fill = "Peak_type", color = "black", palette = "Paired", facet.by = "Peak_type") +
    ylab("Max Height")+
  guides(x =  guide_axis(angle = 45)) +
  ggtitle(paste0(x, " condition"), subtitle = "t-test") +
  stat_compare_means(label = "p.signif", method = "t.test", ref.group = as.character(reference_sample)) 

print(p3)
ggsave(paste0(x, "_barplot_Height_max_statistics.png"), plot = p3, path = "./Figures/Statistics")

p4 <- ggbarplot(df_plot_tumor, "Sample_name", "Height_max_percentage_rep", 
       add = c("mean_sd"),fill = "Peak_type", color = "black", palette = "Paired", facet.by = "Peak_type") +
    ylab("Max Height in %")+
  guides(x =  guide_axis(angle = 45)) +
  ggtitle(paste0(x, " condition"), subtitle = "t-test") +
  stat_compare_means(label = "p.signif", method = "t.test", ref.group = as.character(reference_sample)) 

print(p4)
ggsave(paste0(x, "_barplot_Height_max_percentage_statistics.png"), plot = p3, path = "./Figures/Statistics")



}
```


```{r sessionInfo}
sessionInfo()
```








